DRAFT: Harmonisation and trait aggregation paper

### 2020-12-21 Paper/Draft_harmonization_paper.tex
21.12: Edits Ralf Abstract and References

\section*{Summary}

@@

1. In recent years, comprehensive invertebrate trait databases have been established for different continents. The wide availability of invertebrate trait data should have created an impetus for trait-based studies, especially on large scales. However, a number of data-related problems hamper the work with trait data. For example, standardised trait definitions are lacking and coding of traits varies across trait databases impeding trait data synthesis. Another issue is different taxonomic resolutions between observational taxonomic data and trait databases, leading to the need to aggregate traits. 
% The aim of our study was to investigate discrepancies in trait definitions, to evaluate different trait aggregation methods and

2. In this study, we describe discrepancies in trait definitions of trait databases from Europe, North America, New Zealand, and Australia for traits frequently used in ecological studies. Based on these trait databases we established 4 novel trait datasets by harmonising 7 commonly used grouping features. Two of these datasets were used to evaluate different trait aggregation methods by comparing aggregated traits with assigned traits by experts at family-level. Furthermore, we explored the effect of grouping feature harmonisation and trait aggregation by re-analysing a study on the effect of anthropogenic salinisation on biological traits using harmonised grouping features and aggregated traits.

3. We found that investigated trait databases are distinct with regard to how grouping features are differentiated into traits, the scope of the grouping feature definitions, how traits are defined, and the coding to describe traits. Comparison of aggregated and assigned traits showed that aggregated traits are generally in conclusion with expert assignments, especially approaches using the median. Grouping feature harmonisation and trait aggregation had a minor impact in the re-analysis, although averaging effects can occur due to harmonisation diminishing the signal of a trait.

@@

Explaining and predicting how communities are shaped by environmental factors is one of the main goals of ecology. Species traits, defined as measurable properties of an organism (\cite{mcgill_rebuilding_2006}), might be beneficial in achieving this goal (\cite{heino_jani_macroecological_2013}). Traits evolve through adaptations (e.g., physiological, behavioural, etc.) of organisms to their environment and indicate direct or indirect linkages between the biological response of an organism or a population and its environment (\cite{southwood_habitat_1977, verberk_delivering_2013}).
Besides providing a mechanistic explanation of species-environment relationships, trait-based approaches might be suitable for large scale analysis because the variability in trait responses is lower than in taxonomic responses (\cite{bonada_taxonomic_2007, baird_toward_2011}). Traits of freshwater invertebrate individuals are difficult to determine because - unlike plant traits - they often can not be measured directly. For example, to gain knowledge on feeding habits requires evaluating mouthpart morphology, consumed food, and the organisms function within its community (\cite{moog_comprehensive_nodate}). Nevertheless, invertebrate traits have been increasingly used in freshwater ecology, e.g. by relating macroinvertebrate trait composition to environmental factors and as trait metrics in biomonitoring (\cite{poff_developing_2010, szocs_effects_2014, bhowmik_large_2015, menezes_beyond_2010}).

>>>>

\section*{Abstract}

@@

1. Invertebrate traits are expected to be suitable for large-scale analysis in freshwater ecology due to their consistent response to environmental gradients. In recent years, comprehensive invertebrate trait databases have been established at different scales (e.g. regions, continents). The wide availability of invertebrate trait data supports trait-based studies, especially on large scales. However, a number of data-related problems hamper the work with trait data. For example, for freshwater invertebrate traits standardised definitions are lacking and coding of traits varies across trait databases impeding trait data synthesis. Moreover, uncertainty remains how harmonising grouping features from different trait databases can influence trait-environment relationships. Another issue is that if the taxonomic resolution of observational taxonomic data is less precise than of trait databases, the traits in databases require aggregation before they can be linked to the observational data.
% Here: Describe aims of the study or redundant with part 2? Something like this: The aim of our study is to investigate these data-related issues by synthesizing trait data from different databases.

2. We describe discrepancies in trait definitions of trait databases from Europe, North America, New Zealand, and Australia for traits frequently used in ecological studies. Based on these trait databases, we established 4 novel trait datasets by harmonising 7 commonly used grouping features. Two of these datasets were used to evaluate different trait aggregation methods by comparing aggregated traits with assigned traits by experts at family-level. We tested trait aggregation based on the mean and based on the median, each with different weightings. Furthermore, we explored the effect of grouping feature harmonisation and trait aggregation by re-analysing data from a case study.

3. Between the trait databases, grouping features varied in terms of their classification into traits, the scope of the grouping feature definitions, how traits are defined, and the coding to describe traits. Aggregated and assigned traits based on expert knowledge displayed a good match, specially approaches using the median. Grouping feature harmonisation and trait aggregation only slightly changed the results in the re-analysis compared to the original study, although averaging effects can occur due to harmonisation.

@@

Explaining and predicting how communities are shaped by environmental factors is one of the main goals of ecology. Species traits, defined as measurable properties of an organism (\cite{mcgill_rebuilding_2006}), might be beneficial in achieving this goal (\cite{heino_jani_macroecological_2013}). Traits evolve through adaptations (e.g., physiological, behavioural, etc.) of organisms to their environment and indicate direct or indirect linkages between the biological response of an organism or a population and its environment (\cite{southwood_habitat_1977, verberk_delivering_2013}).
Besides providing a mechanistic explanation of species-environment relationships, trait-based approaches might be suitable for large-scale analysis because the variability in trait responses is lower than in taxonomic responses (\cite{bonada_taxonomic_2007, baird_toward_2011}). Traits of freshwater invertebrate individuals are difficult to determine because - unlike plant traits - they often can not be measured directly. For example, to gain knowledge on feeding habits requires evaluating mouthpart morphology, consumed food, and the organisms function within its community (\cite{moog_comprehensive_nodate}). Nevertheless, invertebrate traits have been increasingly used in freshwater ecology, e.g. by relating macroinvertebrate trait composition to environmental factors and as trait metrics in biomonitoring (\cite{poff_developing_2010, szocs_effects_2014, bhowmik_large_2015, menezes_beyond_2010}).
